Analysis-ready VCF at Biobank scale using Zarr

### 2024-04-30 paper.tex

\subsection{Zarr and block-based compression}
% Zarr is a simple layer on top of best-in-class, modern components
In the interest of completeness it is useful to provide a high-level overview
of Zarr and the techologies that it depends upon. Zarr is a specialised format
for storing large-scale $n$-dimensional numerical data (arrays). Arrays
are split into chunks, which are compressed and stored separately. Chunks are 
addressed by their indexes along the dimensions of the array, and the 
compressed data associated with this key. Chunks are
are often stored in individual files, but a wide array of different
stores are supported including cloud buckets and NoSQL databases. 
Metadata describing the array and its properties is then stored 
in JSON format along with the chunks. Zarr is best seen as a 
\emph{specification} that describes the details of how chunks are 
addressed and stored: there are multiple implementations across
different languages.

Zarr is flexible in allowing different compression codecs and 
pre-compression filters to be specified on a per-array basis.
Two key technologies used by Zarr are the Blosc
% This seems like the best citation for Blosc? It is mentioned here
meta-compressor~\cite{alted2010modern}
and Zstandard compression algorithm~\citep{collet2021rfc}.
Blosc is a high-performance compressor optimised for numerical
data which uses ``blocking''~\citep{alted2010modern} to 
optimise CPU-cache access patterns, as well as highly optimised
bit and byte shuffle filters.  Remarkably, on highly 
compressible datasets, Blosc decompression can be faster 
than \texttt{memcpy}.
Blosc is written in C, with APIs for C, Python, Julia, Rust
and others.

Blosc is a ``meta-compressor'' because it provides different 
access to several different compression codecs. The 
Zstandard compressor~\citep{collet2021rfc} is of particular 
interest here as it achieves very high compression ratios
with fast decompression speeds [see fig]. 
% This may seem a bit off topic, but want to reassure reader that 
% this isn't some niche technology
Zstandard is also used in several recent VCF compression 
methods~\citep[e.g.][]{lefaive2021sparse,wertenbroek2022xsi}.
% LZ4 is also notable - any citations for this?

% Zarr was originally developed as a simpler version of 
% HDF5~\citep{folk2011overview}

Zarr is now widely used.

Multi terabyte numerical weather prediction models~\citep{gowan2022using}
Petabyte scale geoscience datasets in the Pangeo project

>>>>

\subsection{Zarr and block-based compression}
% Zarr is a simple layer on top of best-in-class, modern components
In the interest of completeness it is useful to provide a high-level overview
of Zarr and the techologies that it depends upon. Zarr is a specialised format
for storing large-scale $n$-dimensional data (arrays). Arrays
are split into chunks, which are compressed and stored separately. Chunks are 
addressed by their indexes along the dimensions of the array, and the 
compressed data associated with this key. Chunks are
are often stored in individual files, but a wide array of different
stores are supported including cloud buckets and NoSQL databases;
in principle, Zarr can store data in any key-value store.
Metadata describing the array and its properties is then stored 
in JSON format along with the chunks. The simplicity and transparency
of this design has substantial advantages over other technologies
such as HDF5~\citep{folk2011overview} which are relatively complex and opaque.
This simplicity has led to numerous implementations of the Zarr specification
being developed, ranging from the mature Zarr-Python~\citep{zarrpython}
and TensorStore~\citep{tensorstore} implementations
to more experimental extensions to packages like
GDAL~\citep{gdal_zarr},
NetCDF~\citep{netcfd_c},
N5~\citep{n5zarr}
and xtensor~\citep{xtensor_zarr}
as well as  standalone libraries for JavaScript~\cite{zarrjs},
Julia~\cite{zarrjl}, Rust~\citep{zarrs}
and R~\cite{pizzarr}.

Zarr is flexible in allowing different compression codecs and 
pre-compression filters to be specified on a per-array basis.
Two key technologies used by Zarr are the Blosc
% This seems like the best citation for Blosc? It is mentioned here
meta-compressor~\cite{alted2010modern}
and Zstandard compression algorithm~\citep{collet2021rfc}.
Blosc is a high-performance compressor optimised for numerical
data which uses ``blocking''~\citep{alted2010modern} to 
optimise CPU-cache access patterns, as well as highly optimised
bit and byte shuffle filters.  Remarkably, on highly 
compressible datasets, Blosc decompression can be faster 
than \texttt{memcpy}.
Blosc is written in C, with APIs for C, Python, Julia, Rust
and others.
Blosc is a ``meta-compressor'' because it provides different 
access to several different compression codecs. The 
Zstandard compressor~\citep{collet2021rfc} is of particular 
interest here as it achieves very high compression ratios
with fast decompression speeds [see fig]. 
% This may seem a bit off topic, but want to reassure reader that 
% this isn't some niche technology
Zstandard is also used in several recent VCF compression 
methods~\citep[e.g.][]{lefaive2021sparse,wertenbroek2022xsi}.
% LZ4 is also notable - any citations for this?

This combination of Zarr's simple and cloud-friendly storage 
of data chunks with state-of-the-art compression methods has 
led to Zarr gaining significant traction in large-scale 
scientific computing, in particular in the geosciences.

Multi terabyte numerical weather prediction models~\citep{gowan2022using}

CMIP6 dataset in Google Cloud~\cite{cmip6_dataset}

The Open GeoSpatial consortium has formally recognised Zarr as a community
standard~\cite{ogc_zarr2_standard}
and has formed
a new GeoZarr Standards Working Group to establish a Zarr encoding for
geospatial data~\cite{ogc_geozarr_news}.

Zarr is being considered a potential format for migrating the 
Copernicus global land service to the cloud~\citep{abernathey2021opening}
